\documentclass[dsc,numbers]{coppe}

\usepackage[disable,colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
% \usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\usepackage{xargs}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb}

\usepackage{algorithmic}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}

\usepackage[T1]{fontenc}
\usepackage{textcomp}
\usepackage{listings}

%\usepackage[portuguese]{babel}

\usepackage{hyperref}
\usepackage{imakeidx}
\usepackage{letterspace}
\usepackage{soul}
\usepackage{mfirstuc}

\usepackage{multirow}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{supertabular}
\usepackage{float}
\usepackage{xcolor,colortbl}
\usepackage{makecell}

\usepackage[justification=centering]{caption}

%\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{amsthm}

\usepackage{siunitx}
\sisetup{output-exponent-marker=\ensuremath{\mathrm{e}}}

\MFUnocap{in}
\MFUnocap{or}
\MFUnocap{of}

\usepackage{caption}
\usepackage{subcaption}

\include{macros}

\makeindex
\makelosymbols
\makeloabbreviations


\begin{document}


  \title{\thesistitle{}}
  \foreigntitle{\thesistitlept}
  \author{Renan Francisco Santos}{Souza}
  \advisor{Prof\textsuperscript{a}.}{Marta Lima de Queir\'os}{Mattoso}{D.Sc.}
  \advisor{Prof.}{Patrick}{Valduriez}{Dr.}

  \examiner{Prof\textsuperscript{a}.}{Marta Lima de Queir\'os Mattoso}{D.Sc.}
  \examiner{Prof.}{Patrick Valduriez}{Dr.}
  \examiner{Dr.}{Marco Aurelio Stelmar Netto}{Ph.D.}
  \examiner{Prof.}{Alexandre de Assis Bento Lima}{D.Sc.}
  \examiner{Prof\textsuperscript{a}.}{Vanessa Braganholo Murta}{D.Sc.}
  \examiner{Prof\textsuperscript{a}.}{Lúcia Maria de Assumpção Drummond}{D.Sc.}

  \department{PESC}

  \date{12}{2019}

  \keyword{User Steering}
  \keyword{Data Provenance}
  \keyword{Large-scale Workflows}

  % Do not comment these two below
  \maketitle
  \frontmatter



   %\include{chapters/extras/dedic}
  \include{chapters/extras/thanks}
  \include{chapters/extras/resumo}
  \include{chapters/extras/abstract}
  \tableofcontents
  \listoffigures
  \listoftables
  %\printlosymbols % glossary
  \printloabbreviations

   % Do not comment mainmatter
   \mainmatter



  \include{chapters/chap01}
  \include{chapters/chap02}
  \include{chapters/chap03}
 \include{chapters/chap04}

  \include{chapters/chap_DfAdapter}
  \include{chapters/chap_dChiron}

  \include{chapters/experiments/index}
  \include{chapters/chap07}




  \backmatter
  \bibliographystyle{coppe-unsrt}
  \bibliography{thesis}

  \appendix
  \include{appendix/sql}
  \include{appendix/dfadapter}
  
 

\end{document}
